---
title: "organized"
author: "Theo-60985751"
date: "2023-12-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(pheatmap)
library("TCGAbiolinks")
library("survival")
library("survminer")
library("SummarizedExperiment")
library(DESeq2)
library("gridExtra")
library("AnnotationDbi")
library("org.Hs.eg.db")
library(pathview)
library(gage)
```

# Opening the datafiles
```{r}
data_mutation = read.delim("data_mutations.txt", header = TRUE, sep = '\t')
data_clinical = read.table("data_clinical_patient.txt",header = TRUE, sep = '\t')
data_rna = read.csv("RNAseq_LIHC.csv", header = TRUE, row.names = "X")
```

# Finding the patients that we have data for clinical, genome, and RNAseq data
```{r}
unique_patient_clinical = unique(data_clinical$PATIENT_ID)
unique_patient_rna = unique(colnames(data_rna))
unique_patient_mutation = unique(data_mutation$Tumor_Sample_Barcode)

shortened_rna = substr(unique_patient_rna, start = 1, stop = 12)
shortened_rna = gsub("\\.", "-", shortened_rna)

shortened_mutation = substr(unique_patient_mutation, start = 1, stop = 12)

common_names1 <- intersect(unique_patient_clinical, shortened_mutation)
common_names2 <- intersect(common_names1, shortened_rna)

```

###  Adding the patient ID at the last column and get the common datas
```{r}
data_clinical_common = subset(data_clinical, PATIENT_ID %in% common_names2)

data_mutation$PATIENT_ID = substr(data_mutation$Tumor_Sample_Barcode, start = 1, stop = 12)
data_mutation_common = subset(data_mutation, PATIENT_ID %in% common_names2)

data_mutation_common$Tumor_Sample_Barcode = data_mutation_common$PATIENT_ID

data_rna_shortened = data_rna
colnames(data_rna_shortened) = substr(colnames(data_rna_shortened), start = 1, stop = 12)
colnames(data_rna_shortened) = gsub("\\.", "-", colnames(data_rna_shortened))
data_rna_common = data_rna_shortened[,common_names2]

```

### Generate a plot for distribution of variant classifications
```{r}
data_oncoplot = data_mutation_common

hugo <- as.data.frame(table(data_oncoplot$Hugo_Symbol))
var.class <- as.data.frame(table(data_oncoplot$Variant_Classification))
ggplot(data=var.class, aes(x=Var1, y=Freq))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 45,hjust=1))
```

### Generate a plot for distribution of variant classifications
```{r}
data_oncoplot = data_mutation_common

hugo <- as.data.frame(table(data_oncoplot$Hugo_Symbol))
var.class <- as.data.frame(table(data_oncoplot$Variant_Classification))
ggplot(data=var.class, aes(x=Var1, y=Freq))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 45,hjust=1))
```
### Generate a plot for distribution of mutation events
```{r}
hugo <- as.data.frame(table(data_mutation_common$Hugo_Symbol))

hugo.ordered <- hugo[order(-hugo$Freq),]

ggplot(data=hugo.ordered[1:20,], aes(x=Var1, y=Freq))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 45,hjust=1))+
  scale_x_discrete(limits = hugo.ordered[1:20,]$Var1)
```
### Generate an oncoplot matrix of all genes
```{r}
cnv_events = unique(data_oncoplot$Variant_Classification)
oncomat = reshape2::dcast(
  data = data_oncoplot,
  formula = Hugo_Symbol ~ Tumor_Sample_Barcode,
  fun.aggregate = function(x, cnv = cnv_events) {
    x = as.character(x) # >= 2 same/distinct variant classification = Multi_Hit
    xad = x[x %in% cnv]
    xvc = x[!x %in% cnv]
    
    if (length(xvc) > 0) {
      xvc = ifelse(test = length(xvc) > 1,
                   yes = 'Multi_Hit',
                   no = xvc)
    }
    
    x = ifelse(
      test = length(xad) > 0,
      yes = paste(xad, xvc, sep = ';'),
      no = xvc
    )
    x = gsub(pattern = ';$',
             replacement = '',
             x = x)
    x = gsub(pattern = '^;',
             replacement = '',
             x = x)
    return(x)
  },
  value.var = 'Variant_Classification',
  fill = '',
  drop = FALSE
)

rownames(oncomat) = oncomat$Hugo_Symbol
oncomat <- oncomat[,-1]

oncomat.ordered <- oncomat[order(-hugo$Freq),]
```

### Transform the matrix into a binary matrix
```{r}
mat <- oncomat.ordered
mat[mat== "Silent"] = 0
mat[mat == "Intron"] = 0
mat[mat == "Missense_Mutation"] = 0
mat[mat == ""] = 0

mat <- apply(mat, 2 ,as.numeric)
mat <- as.matrix(mat)
mat[is.na(mat)]=1

rownames(mat)  <-  row.names(oncomat.ordered)
```
### Finding the top 20 most mutated genes
```{r}
genes = rowSums(mat)
oncomat.ordered <- oncomat[order(-hugo$Freq),]
genes.ordered = sort(genes, decreasing = TRUE)


genes.ordered.top = genes.ordered[1:20]
genes.ordered.top.names = names(genes.ordered.top)

data_oncoplot.top = subset(data_oncoplot, Hugo_Symbol %in% genes.ordered.top.names)

```

### Making a matrix of the top 20 mutated genes
```{r}
cnv_events = unique(data_oncoplot.top$Variant_Classification)
oncomat.top = reshape2::dcast(
  data = data_oncoplot.top,
  formula = Hugo_Symbol ~ Tumor_Sample_Barcode,
  fun.aggregate = function(x, cnv = cnv_events) {
    x = as.character(x) # >= 2 same/distinct variant classification = Multi_Hit
    xad = x[x %in% cnv]
    xvc = x[!x %in% cnv]
    
    if (length(xvc) > 0) {
      xvc = ifelse(test = length(xvc) > 1,
                   yes = 'Multi_Hit',
                   no = xvc)
    }
    
    x = ifelse(
      test = length(xad) > 0,
      yes = paste(xad, xvc, sep = ';'),
      no = xvc
    )
    x = gsub(pattern = ';$',
             replacement = '',
             x = x)
    x = gsub(pattern = '^;',
             replacement = '',
             x = x)
    return(x)
  },
  value.var = 'Variant_Classification',
  fill = '',
  drop = FALSE
)
hugo <- as.data.frame(table(data_oncoplot.top$Hugo_Symbol))

rownames(oncomat.top) = oncomat.top$Hugo_Symbol
oncomat.top <- oncomat.top[,-1]
oncomat.top.ordered <- oncomat.top[order(-hugo$Freq),]
```

### Generate a plot for distribution of the top 20 most mutated genes
```{r}
hugo.ordered <- hugo[order(-hugo$Freq),]
ggplot(hugo.ordered, aes(x=Var1, y=Freq))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 45,hjust=1))+
  scale_x_discrete(limits = hugo.ordered$Var1)
```

### Transforming the top matrix into binary
```{r, echo = FALSE}
mat.top <- oncomat.top.ordered
mat.top[mat.top== "Silent"] = 0
mat.top[mat.top == "Intron"] = 0
mat.top[mat.top == "Missense_Mutation"] = 0

mat.top[mat.top == ""] = 0


mat.top <- apply(mat.top, 2 ,as.numeric)
mat.top <- as.matrix(mat.top)
mat.top[is.na(mat.top)]=1

rownames(mat.top)  <-  row.names(oncomat.top.ordered)
```
### Generate a pheatmap
```{r}
reduce.mat <- mat.top[1:3,]
res <- pheatmap(reduce.mat,
         cluster_rows = F,
         show_colnames = FALSE)
```
### clustering
```{r}
cluster_mutation = as.data.frame(cutree(res$tree_col, k = 2))
cluster_mutation
```
### Finding the patientID in each group
```{r}
mutation_group1_patientID = subset(cluster,`cutree(res$tree_col, k = 2)` == 1)
mutation_group2_patientID = subset(cluster,`cutree(res$tree_col, k = 2)` == 2)

data_clinical_common_group1 = subset(data_clinical_common,PATIENT_ID %in% rownames(mutation_group1_patientID))

data_clinical_common_group2 = subset(data_clinical_common,PATIENT_ID %in% rownames(mutation_group2_patientID))

```


### survival analysis on clinical data 
```{r}
data_clinical_common$deceased = data_clinical_common$PFS_STATUS == "1:PROGRESSION"

# create an "overall survival" variable that is equal to days_to_death
# for dead patients, and to days_to_last_follow_up for patients who
# are still alive
data_clinical_common$overall_survival = ifelse(data_clinical_common$deceased,
                                   data_clinical_common$OS_MONTHS,
                                   data_clinical_common$DAYS_LAST_FOLLOWUP)
```

### SA by cluster

```{r}
# Adding a cluster annotation column

data_clinical_common$cluster = cluster_mutation[data_clinical_common$PATIENT_ID,]

Surv(data_clinical_common$overall_survival, data_clinical_common$deceased) ~ data_clinical_common$cluster

table(data_clinical_common$cluster)
fit = survfit(Surv(overall_survival, deceased) ~ cluster, data = data_clinical_common)

pval = surv_pvalue(fit, data=data_clinical_common)$pval
print(pval)

ggsurvplot(fit, data=data_clinical_common, pval=T, risk.table=T, risk.table.col="strata", risk.table.height=0.35)
```

# Performing clustering on RNA-seq data

```{r}
data_rna_common <- data_rna_common[rowSums(data_rna_common)>1,]
sampleDists = dist(t(data_rna_common),upper = TRUE)

```
```{r}
sampleDistMatrix = as.matrix( sampleDists )
rownames(sampleDistMatrix) = colnames(data_rna_common)
colnames(sampleDistMatrix) = colnames(data_rna_common)

res_rna = pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE)
```


```{r}
cluster_rna = as.data.frame(cutree(res_rna$tree_col, k = 2))
cluster_rna
```

Running DE on the clusters
```{r}
rownames(cluster_rna) = substr(rownames(cluster_rna), start = 1, stop = 12)
common_cluster <- row.names(cluster_rna)

data_rna_common_cluster <- data_rna_common[,common_cluster]

metadata_rna <- data.frame(
  patientID = row.names(cluster_rna),
  condition = cluster_rna$`cutree(res_rna$tree_col, k = 2)`
)

### making metadata a factor so DESeq2 runs faster
row.names(metadata_rna) = metadata_rna$patientID
metadata$patientID = NULL
metadata$condition = factor(metadata$condition)
```

```{r}
dds_rna <- DESeqDataSetFromMatrix(countData = data_rna_common_cluster, colData = metadata, design = ~condition)
```


```{r}
dds_rna = DESeq(dds_rna)
dds_rna
```


Normalizing the data
```{r}
# this gives log2(n + 1)
ntd <- normTransform(dds_rna)
# Variance stabilizing transformation
vsd <- vst(dds_rna)

# Regularized log transformation
# The blind=TRUE argument results in a transformation unbiased to sample condition information.
# rld <- vst(dds, blind=FALSE)
```

```{r}
res_de_rna = results(dds_rna, contrast=c(1, 2))
mcols(res_de_rna, use.names = TRUE)
summary(res_de_rna)
res_de_rna.05 <- results(dds_rna, alpha = 0.05)
table(res_de_rna.05$padj < 0.05)

res_rnaLFC1 <- results(dds_rna, lfcThreshold=1)
table(res_rnaLFC1$padj < 0.1)
```


```{r}
# Top20genes = head(resSig[order(resSig$log2FoldChange, decreasing=TRUE),], n= 20)

# genes <- order(res$padj, decreasing = TRUE) [1:20]

# we can select a subset of genes to plot.let’s choose the 20 genes with the largest positive log2fold change.
genes <- order(res_de_rna$log2FoldChange,decreasing = TRUE)[1:20]

# or largest negative log2fold change
# genes <- order(res$log2FoldChange, decreasing = FALSE)[1:20]

# or we can select the top 20 significant genes
```



